---
title: "cont_cluster"
output: pdf_document
date: "2024-12-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
At first glance, there doesn't seem to be any obvious grouping by geography.
Group 1 seems like it is the teams that did better in the tournament.
Let's evaluate more carefully.

Let's split the countries into 5 groups based on geography.
Europe and Africa can be their own groups.
Forming the other 3 groups is more interesting.
Let's try a group of countries from the Americas, not including Canada and the 
USA. Let's make another group for Asian countries. The last group can be 
Canada, USA, and Australia. The reasoning here is that Canada, USA, and Australia
seem more culturally similar to each other and Mexico, central America, and South
America seem more culturally similar to each other. Objectively, this is at least
true in terms of language.



```{r}
coords_df <- read.csv("avg_team_data_kmeans.csv")
continent_g1 <- c("latam", "can_us_aus", "eur", "latam", "afr", "can_us_aus",
                  "latam", "eur", "eur", "latam", "eur", "eur", "eur", "afr",
                  "asia", "asia", "asia", "latam", "afr", "eur", "eur", "eur", 
                  "asia", "asia", "afr", "eur", "eur", "eur", "afr", "can_us_aus",
                  "latam", "eur")

coords_df$cont_g1 <- continent_g1
write.csv(coords_df, "avg_team_data_cont.csv")
