---
title: "kmeans"
output: pdf_document
date: "2024-12-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Does geography influence play style? 

That is, can we cluster the data such that the clusters represent distinct geographical regions?

```{r}
library(ggplot2)
coords_df <- read.csv(
  "avg_team_data.csv",
  header=TRUE
)

#Remove variables not to be included in kmeans
kmeans_df <- coords_df[,-c(1:6,ncol(coords_df))]
#Scale data
kmeans_df <- scale(kmeans_df)


#Find k that minimizes within sum of squares (wss)

#Store wss for each k value
wss <- list()
for (i in 1:10) {
  # Fit the model: km.out
  kmeans_wss <- kmeans(kmeans_df, centers = i)$tot.withinss
  # Save the within cluster sum of squares
  wss[[i]] <- kmeans_wss
  
  
  
}


#Scree plot
scree_df <-  data.frame(wss=unlist(wss), k=1:10)

kmeans_scree <- ggplot(scree_df, aes(x = k, y = wss)) +
    geom_point()+
    geom_line() +
    xlab('K')
kmeans_scree
ggsave("kmeans_scree.png")

```
Looks like the rate of decrease drops off after k=5.
Try clustering with k=5.
This seems promising given that it is close to the number of continents.
We can creatively group countries together to make 5 groups

```{r}
set.seed(42)
k5 <- kmeans(kmeans_df, centers = 5, nstart = 20)
k5$size
#Add clusters to df
coords_df$cluster5 <- as.factor(k5$cluster)
write.csv(coords_df, "avg_team_data_kmeans.csv")

```
There is a group of one!
Let's see which country did not fit into any group.
Any guesses?