---
title: "linreg"
output: pdf_document
date: "2024-12-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Comparison to linear model

```{r}
library(caret)
train_data_reg <- read.csv("xgb_train.csv")
test_data_reg <- read.csv("xgb_test.csv")
test_y <- read.csv("xgb_test_y.csv")
df <- read.csv("match_data_clean.csv")
#Linear regression model
lin_model <- lm(total.goals~., data=as.data.frame(train_data_reg))
#Predictions
lin_pred <- predict(lin_model, test_data_reg)
write.csv(lin_pred, "lin_pred.csv")
#Evaluate performance
#MSE
mean((test_y - lin_pred)^2) 
#MAE
MAE(test_y, lin_pred) 
#RMSE
RMSE(test_y, lin_pred)

mean(df$total.goals)
```
As suspected, the model's performance is poor. All three of the evaluation 
metrics are horrendous considering that the mean number of total goals is 2.6875.
Furthermore, the fit is rank-deficient since there are more predictors than
observations. This results in very poor predictions.
This example illustrates the importance of reducing multicollinearity and variable
selection.
