---
title: "pfi_df"
output: pdf_document
date: "2024-12-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
pfi_df <- read.csv("pfi_df.csv")
# Plot the coefficients for each class
ggplot(pfi_df, aes(x=Var, y=PFI)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "PFI Values", 
       x = "Feature", 
       y = "PFI") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=6)) 
ggsave("pfi_plot.png")
```


A negative PFI values indicates that the model accuracy decreased when permuting that particular variable, and vice versa for positive PFI values. A PFI
value of 0 indicates that there was no change in model accuracy.

The PFI results differed from the coefficients values. Notably, the PFI for red
cards was 0. Forced turnovers was the second lowest PFI value, even though it did not have a particularly large coefficient value. Total 
attempts on goal was the most influential variable in terms of PFI and in terms
of coefficient values
