---
title: "possession_plot"
output: pdf_document
date: "2024-12-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory visualizations

```{r}
library(ggplot2)
df <- read.csv("match_data_clean.csv", header=TRUE)
#Is the possession split more evenly in elimination games than in group games?
pos_plot <- ggplot(df, aes(x = possession.team1, y = possession.team2, color = elimination)) +
  geom_point(size = 3) +
  labs(x = "Possession team 1", y = "Possession team 2", color = "Elimination Game") +
  ggtitle("Possession comparison coloured by elimination game indicator") +
  scale_x_continuous(limits = c(10, 80)) +
  scale_y_continuous(limits = c(10, 80))


ggsave("pos_plot.png")




```

The distribution of the total number of goals is skewed to the right.
There appears to be a positive correlation between total attempts and total
attempted line breaks.
Games that feature more total attempts and total attempted line breaks tend
to have a greater number of goals scored. You can visualize this by selecting
groups of points in the scatter plot and seeing how the barplot bars fill up.
There does not appear to be any significant difference in possession split when
comparing elimination vs group stage games.